### 2020-02-07 paper/sex_selection_spacing-ver3.tex
Make paper more concise
- Rearrange sections
- Rewrite fertility section
- Rewrite mortality section
- Move/combine graphs

\usepackage{endfloat}

@@

India has changed significantly over the last four decades:
the economy has grown substantially,
education levels have increased for both males and females,
and the total fertility rate has fallen to 2.2, or just above replacement 
\citep{Bosworth2008,Dharmalingam2014,
International-Institute-for-Population-Sciences-IIPS2017}.

India has, however, also experienced a rapidly expanding 
access to prenatal sex determination.
Combined with continued strong son preference, the result has been a dramatic 
increase in the males-to-females ratio at birth
\citep{das_gupta97,Sudha1999,Arnold2002,retherford03b,Guilmoto2012,
Bongaarts2013,Portner2015b,Jayachandran2017}.

What the prior research on India has not appreciated is that since each abortion 
increases the interval between births by six to twelve months, the growing use of 
sex-selective abortions has the potential to substantially increase birth spacing.%
\footnote{
The increase consists of three parts.
First, starting from the abortion, the uterus needs at 
least two menstrual cycles to recover;  otherwise, the likelihood 
of spontaneous abortion increases substantially \citep{zhou00b}.
The second part is the waiting time to conception, which is between 
one and six months \citep{Wang2003}.
Finally, sex determination tests are reliable only from three months 
of gestation onwards.
}
Furthermore, the combination of increasing female education, 
lower fertility, higher household income, and the low and declining female 
labor force participation are also likely to impact birth spacing.
Hence, the changes in birth spacing may be much more substantial in India 
than what we have observed in other countries as they developed.

Understanding how birth spacing changes is essential for three reasons.

First, tempo effects, such as those arising from changes in birth spacing, bias 
the total fertility rate (TFR), in the sense that the TFR becomes a more biased measure 
of cohort fertility \citep{Hotz1997,Bongaarts1999,Ni-Bhrolchain2011}.
If birth spacing increased substantially over time, India's fertility may, in 
fact, be higher than generally accepted based on the TFR.

Second, longer birth spacing is associated with lower mortality and 
morbidity risk \citep{Conde-Agudelo2012,Molitoris2019}.
In India, the duration to the next birth has traditionally been shorter with fewer sons, 
which likely contributed to the higher mortality and morbidity risk for girls
\citep{Whitworth2002,Bhargava2003,Bhalotra2008,Maitra2008,
Jayachandran2011,Jayachandran2017a}.
Longer spacing between births then potentially lowers mortality, although there may
be counteracting effects if the longer spacing arises because of multiple abortions, but
there has so far been no attempt to examine which mechanism dominates.

Finally, birth spacing has long served as a measure of son preference, with strong
son preference associated with shorter birth intervals, the fewer sons a family has 
\citep{ben-porath76b,Leung1988}.
Now, longer birth intervals in the absence of sons may instead arise because couples 
with a strong preference for sons use sex selection, possibly to the point where we 
observe a reversal of the traditional spacing pattern.
Despite a large literature on son preference and birth spacing, there is, however,
no prior research has examined how sex selection affects spacing.

Here, I examine how birth spacing in India has changed over time and across 
groups with the introduction of sex selection.
Using data from the four National Family and Health Surveys (NFHS), I apply
a competing risk hazard model to the birth histories of Hindu women, covering
the period 1972 to 2016.
The competing risk hazard model directly incorporates the effects of 
sex-selective abortions on birth intervals
\emph{and} 
the likelihood of a son, 
but also works in cases without prenatal sex selection.

@@

Third, to what extent does TFR reflect cohort fertility?
An advantage of the empirical model is that I can predict fertility behavior taking into
account both the likelihood of parity progression and having a child of a specific sex.
I compare those predictions with fertility rate estimates based on the same data to
show how the increases in birth spacing have led to a substantial overestimate of how 
fast fertility has declined. [TK too imprecise? underestimate of the level, rather?]

@@

To set the stage for the subsequent analyses, I first describe how female education
and labor force participation has changed.
Both are potentially important factors in the birth spacing decision, as I discuss
in the subsequent section.

@@

Female education is higher in urban areas than in rural areas.
Around sixty percent of urban women born in the 1930s had no education, 25 percent had
between one and seven years, about 10 had eight or more years of education, and only
five percent has twelve years or more.
The proportions with no education and one to seven years have both declined to just below 
ten percent for the latest cohort.
Although the proportion of women with eight to eleven years of education has increased
to about twenty, most of the increase in urban female education has come from the 
twelve plus group, which now account for more than half of all urban women.

@@

There is a U-shaped relationship between education and working in most cases with 
the highest percent working for women with either no education or with twelve or 
more years and the lowest for women with eight to eleven years of education.
Women are more likely to report working if they live in rural than urban areas 
and the older they are.
Comparing 1992 and 2015, the percent of women who currently work has either 
remained roughly the same or decreased.
The cases were the numbers were higher in 1999 and 2006 than in 1992 or 2015 
are mostly based on small samples, such as rural women with twelve or more 
years of education in rural areas.

A question is whether the rapid expansion in female educational attainment has led
to significant changes in the composition of better-educate women and whether this will
affect their son preference and desired fertility.
It is clear that the increases in female educational attainment imply that access to 
education has expanded beyond the higher castes. 
However, ``Sanskritization'' implies that as lower castes females gain access to 
education and their husbands' income increases, the women adopt higher caste 
norms such as stronger son preference and a retraction from the formal labor 
market \citep{Srinivas1956,Chen1995,Abraham2013,Chatterjee2018}.

@@

The standard economic argument for why increasing female education has predominantly
been associated with shorter spacing is that parents incur time costs when they have 
children \citep{Hotz1997,schultz97}.
Specifically, if having children require the mother to reduce her market work, parents 
can lower the cost of children by shorten birth spacing.
With the shorter birth spacing, parents can take advantage of economies of scale in 
childrearing, because taking care of two young children does not take twice as much 
time as taking care of one child \citep{Vijverberg1982}.
In this case, better-educated women will both have lower fertility and shorter
spacing than less-educated women because of their higher opportunity cost of time 
\citep{Ross1974,Newman1984}.

The low and declining female labor force participation, especially for younger women, 
suggests, however, that families face little incentive to space children more closely 
together for economic reasons.
This may happen if household income has increased so substantially that the income effect 
dominates any substitution effect.
Two findings speak for this effect.
First, although real wages for both men and women have almost doubled between 1987 and 
2011, the mean male wage is still close to 70 higher than the female wage 
\citep{Klasen2015,Bhargava2018}.
Second, women’s labor supply appear to be more negatively elastic to their husbands' 
wages than it is positively elastic to own wages \citep{Bhargava2018}.

In fact, it may be that the combination of rising incomes and continued son preference 
lead to even longer spacing than what we would expect from the standard income effect.
As women's education increases, their productivity at home also increases, especially in 
the production of offspring human capital.
With relatively more boys born because of increased access to sex-selective 
abortions and an increasing potential income for (male) offspring, demand for 
better-educated women can increase, even if they do not participate in the labor market 
\citep{Behrman1999}.
The result would be longer birth intervals because that increases parental attention 
per child resulting in higher child ``quality'' \citep{Zajonc1975,Zajonc1976,Razin1980}.
The evidence on spacing's effect on child quality measures such as IQ 
and education is, however, mixed for developed countries and non-existing 
for developing countries
\citep{Powell1993,Pettersson-Lidbom2009,Buckles2012,Barclay2017}.
Nevertheless, for one crucial aspect of child quality---health and mortality---longer 
spacing does lead to better outcomes, although this relationship weakens with maternal 
education \citep{Whitworth2002,Conde-Agudelo2012,Molitoris2019}.

While one might expect that increasing income and female education would also lengthen
average birth spacing through better access and knowledge of contraception, the empirical 
relationship between contraception use and birth spacing is ambiguous 
\citep{Tulasidhar1993,Whitworth2002,Bhalotra2008,Yeakey2009,Kim2010,Soest2018}.
Furthermore, the ability to successfully use ``low efficacy'' contraceptive methods, 
such as the rhythm method, is increasing in education, while there is no difference 
across education levels when it comes to modern contraceptives such as pills or 
injectables \citep{Rosenzweig1989}.

With the introduction of sex selection, parents can avoid the birth of girls, 
although at the cost of the procedure and having to wait longer until the next child 
if each aborted pregnancy.
Theory suggests that sex selection increases with lower desired fertility, and, for a 
given desired number of children, the higher the parity \citep{Portner2015b}.
The increased use of sex selection in India with education and in urban
areas is consistent with a lower desired fertility for both groups
\citep{das_gupta97,retherford03b,Guilmoto2009a,Portner2015b,Jayachandran2017}.
Better-educated women also tend to live in households with higher income, which better 
enables them to access prenatal sex determination and lowers the relative costs of using 
sex selection and have long birth intervals.

In summary, with substantial increases in husband income and a declining female labor 
force participation, I expect a push toward longer birth spacing over time, independent
of education levels, based on the income effects and the effects of spacing
on child outcomes.
As desired fertility decreases with increasing education, I, furthermore, expect 
birth spacing to increase the most among the best-educated because of their 
use of sex selection and because of with "Sanskritization" the changing composition 
of women in the better-educated group does not substantially change this 
group's use of sex selection.%
\footnote{
Working in the opposite direction is that women with more education can space
their children closer together without substantially increasing child mortality risk,
although compared with those factors this effect is likely to be relatively small.
}

@@

The standard approach in the birth spacing literature is to use proportional hazard
models with a single exit, the birth of a child.%
\footnote{
See \citet{Sheps1970} and \citet{Newman1984} for early discussions of why 
hazard models are the preferred way to deal with the censoring of birth
intervals.
}
There are, however, two problems with the standard approach in this setting.
First, and most importantly, the introduction of sex selection means that the sex of the 
next child is no longer necessarily random and that parents' choices will impact the 
spacing to the birth of a girl or a boy differently.
I, therefore, use a competing risk set-up, which can capture both the non-randomness of
the birth outcome and the differential spacing.%
\footnote{
\cite{Merli2000} used a discrete hazard model to examine whether 
there were under-reporting of births in rural China, although they 
estimated separate waiting time regressions for boys and girls.
}

@@

The distribution of spacing is captured by the survival curve, which shows the probability 
of not having had a birth yet by spell duration, for a given set of explanatory variables.
The survival curve at time $t$ is 
\begin{equation}
\label{eq:survival}
S_{t} 
= 
\prod_{d=1}^t
\left(
\frac{ 1 }
{1 + \sum_{l=2}^2 \exp(D_j(t) + \alpha_{ld}'\mathbf{Z}_{kd} + \beta_l'\mathbf{X}_{k})}
\right).
\end{equation}

Because the probability of ever having a next birth varies across groups, a direct 
comparison of standard survival curves tells us little about how the spread of sex 
selection affects birth spacing across groups.
I, therefore, condition on the predicted likelihood of parity progression when examining 
birth spacing measures, such as the average duration to a birth.
The reliability of this approach depends on whether the spell length covered is 
sufficiently long that few women are likely to give birth after the spell cut-off.
I discuss the choice of spell length below.%
\footnote{
It is important to note that the approach is not the same as merely calculating 
the birth spacing measures for women who already have a given 
parity child in the survey because that number does not take into account
the censoring of spells that will eventually lead to a birth.
}
In addition, I present graphs of survival curves conditional on parity progression,
which therefore begin at 100\% and end at 0\%.

Interpretation of the model coefficients is challenging \citep{thomas96}.
It is, however, possible to calculate the predicted probabilities of 
having a boy, $b$, and of having a girl, $g$, in period $t$, conditional on 
a set of explanatory variables and not having had a child before that period, as
\begin{align}
P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt}, t ) 
& =  
\frac{ \exp(D_j(t) + \alpha_{1t}' \mathbf{Z}_{kt} + \beta_1' \mathbf{X}_{k} )}
{1 + \sum_{l=1}^2 \exp(D_j(t) + \alpha_{lt} ' \mathbf{Z}_{kt} + \beta_l ' \mathbf{X}_{k})}
\label{eq:probability_boy} \\
P(g_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t ) 
& =  
\frac{ \exp(D_j(t) + \alpha_{2t}'\mathbf{Z}_{kt} + \beta_2'\mathbf{X}_{k} )}
{1 + \sum_{l=2}^2 \exp(D_j(t) + \alpha_{lt}'\mathbf{Z}_{kt} + \beta_l'\mathbf{X}_{k})}
\label{eq:probability_girl}
\end{align}
It is then straightforward to calculate the estimated percentage of children born that 
are boys, $\hat{Y}$, at each $t$:  
\begin{equation}
\hat{Y}_t 
= 
\frac{ P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )}
{ P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t) + P(g_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )} 
\times 100.
\label{eq:probability_son}
\end{equation}
Combining the percentage boys and the likelihood of exiting the spell 
across all $t$ gives the predicted percent boys born over the entire spell.

@@

In addition to the large number of women surveyed and the long period covered, one of the
benefits of the NFHS is that surveys enumerators pay careful attention to spacing between 
births and probe for ``missed'' births.
Nevertheless, systematic recall error, where the likelihood of reporting a deceased 
child depends on the sex of the child, is still a potentially substantial problem and
will bias both spacing and sex ratios.
As I show in the online Appendix, recall error is heavily dependent on how long ago a 
woman was married, and I, therefore, drop women married 22 years or more.
The final sample consists of 
\input{../tables/num_women.tex}
% [number] women, with [number] parity one through four births.

Direct information on the use of sex selection is not available, so I compare different 
periods as a proxy, based on the changes in access and legality of prenatal sex 
determination in India.
Abortion has been legal in India since 1971.
Reports of sex determination appeared around 1982--83, and the number of clinics 
quickly increased \citep{Sudha1999,bhat06,Grover2006}.
In 1994, the Prenatal Diagnostic Techniques (PNDT) Act made determining and communicating 
the sex of a fetus illegal.%
\footnote{
Details about the act are at \href{http://pndt.gov.in/}{http://pndt.gov.in/}.
There is little evidence that the ban significantly affected sex ratios \citep{Das-Gupta2016}.
}
Although the use of sex selection increased even after 1994, some researchers argue that 
we have passed a turning point in its use \citep{Das_Gupta2009,Diamond-Smith2015}.
I, therefore, use four periods: 1972--1984, 1985--1994, 1995--2004, and 2005--2016.
The allocation of spells into periods is determined by when conception, and therefore 
decisions on sex selection, can begin, even if we do not observe any births until nine 
months later. 
This allocation mechanism means that some spells cover two periods, which may bias 
downward the differences between the periods.

I focus here on the second through fourth spells, that is, on the intervals from the first 
birth until at most the fourth birth.
I exclude the spell from marriage to the first birth because of data issues.
In NFHS-1 and NFHS-2 the only information on marriage is the age at marriage, and 
all first spell durations are therefore imputed.
In NFHS-3 about a third of respondent did not provide complete information on the
timing of marriage and had the length of their first spell imputed using the age at 
marriage.
The imputation results in large numbers of women with spells that are either negative or 
less than nine months and makes a comparison of spell duration over time subject to an 
excessive amount of noise.

The spells all begin nine months after the previous birth, which is the earliest we should 
expect to observe a new birth.
A spell continues until either a child is born or censoring occurs.
Censoring can happen for three reasons:
the survey takes place;
sterilization of the woman or her husband;
or too few births are observed for the method to work.
For all spells censoring is set at 96 months (eight years) after a birth can occur.
With these cut-offs, less than one percent of observed births occur after
the spell cut-off.%
\footnote{
\input{../tables/num_missed.tex}
}

@@

Women with different education levels have different hazard profiles, although the size 
and direction of the effect vary across areas and time 
\citep{Tulasidhar1993,Whitworth2002,Bhalotra2008,Kim2010,Soest2018}.
Furthermore, the use of sex selection increases with education, either because of the 
associated lower fertility or because higher income enables them to access and
afford prenatal sex determination
\citep{das_gupta97,retherford03b,Guilmoto2009a,Bongaarts2013,Portner2015b,Jayachandran2017}.%
\footnote{
It is also possible that the effect is driven by a stronger son preference, although 
higher education women have shown declining son preference \citep{bhat03,pande07}.
}
I divide women into three groups based on education attainment: no
education, one to seven years of education, and eight and more years of education.
The models are estimated separately for each education level to reduce the potential 
problem of including other variables as proportional.

@@

\section{Results\label{sec:results}}

% [How to read tables]
I estimate the model for each spell/education/period subsample, the results of which I use 
to predict average birth spacing, sex ratio, and the probability of having a birth in that 
spell.
Figures \ref{fig:bootstrap_low}, \ref{fig:bootstrap_med}, and  \ref{fig:bootstrap_high} 
show these predicted outcomes for the three education levels separated by the area of 
residence.%
\footnote{
For legibility, none of the graphs show standard errors, which, together with the graphed 
values, are shown in Appendix Tables 
\ref{tab:avg_sex_ratio_low}, 
\ref{tab:avg_sex_ratio_med}, and 
\ref{tab:avg_sex_ratio_high}.
}
The predicted probability of having a birth by the end of the spell (parity progression
probability) is $1-S_{T}$, that is the likelihood of having had a birth by 96 months after 
the beginning of the spell.
To find the expected average duration I first calculate for each woman the probability 
of giving birth in each time $t$, which I use as weights to calculated her expected duration. 
I then average the individual expected durations across women using the parity progression
probabilities as weights.

The predicted sex ratio captures the percent boys that will have been born to women 
in the sample when childbearing for that spell is over.
A woman's predicted sex ratio is the weighted average of the predicted percentage boys 
over each $t$ in the spell, calculated using equation (\ref{eq:probability_son}), using
the probability of giving birth at time $t$ as weights.%
\footnote{
If T=2 with the estimated percent boys 54\% and 66\% and the likelihoods of giving 
birth 20\% and 40\%, the percent boys is $\frac{54*0.2+66*0.4}{0.2+0.4} = 62$.
}
The predicted sex ratio shown is the weighted average of the individual predicted sex 
ratios, using the parity progression probabilities as weights.
For comparison, the graphs also show the natural sex ratio, approximately 51.2\% 
\citep{ben-porath76b,jacobsen99,Portner2015b}.

@@

% \captionsetup[subfigure]{position=top,captionskip=-1pt,farskip=-0.5pt}
\captionsetup[subfigure]{captionskip=-1pt,farskip=-0.5pt}

>>>>

% \usepackage{endfloat}

@@

India has changed significantly over the last four decades:
the economy has grown substantially,
education levels have increased for both males and females,
and the total fertility rate has fallen to 2.2 
\citep{Bosworth2008,Dharmalingam2014,
International-Institute-for-Population-Sciences-IIPS2017}.

India has, however, also experienced a rapidly expanding 
access to prenatal sex determination.
Combined with continued strong son preference, the result has been a dramatic 
increase in the males-to-females ratio at birth
\citep{das_gupta97,Arnold2002,retherford03b,Guilmoto2012,
Portner2015b,Jayachandran2017}.

What the prior research on India has not appreciated is that since each abortion 
increases the interval between births by six to twelve months, the growing use of 
sex-selective abortions may substantially increase birth spacing.%
\footnote{
The increase consists of three parts.
First, starting from the abortion, the uterus needs at 
least two menstrual cycles to recover;  otherwise, the likelihood 
of spontaneous abortion increases substantially \citep{zhou00b}.
The second part is the waiting time to conception, which is between 
one and six months \citep{Wang2003}.
Finally, sex determination tests are reliable only from three months 
of gestation onwards.
}
Furthermore, the combination of increasing female education, 
lower fertility, higher household income, and the low and declining female 
labor force participation are also likely to impact birth spacing.
Hence, the changes in birth spacing may be much more substantial in India 
than what we have observed in other countries.

Understanding how birth spacing changes is essential for two main reasons.

First, if birth intervals increase substantially over time, India's fertility may 
be higher than generally accepted based on the total fertility rate (TFR).
Increasing birth spacing is a tempo effect, which, by postponing births, makes the TFR a 
downward biased estimate of cohort fertility 
\citep{Hotz1997,Bongaarts1999,Ni-Bhrolchain2011}.

Second, longer birth spacing is associated with lower mortality and 
morbidity risk \citep{Conde-Agudelo2012,Molitoris2019}.
In India, the duration to the next birth has traditionally been shorter with fewer sons, 
which likely contributed to the higher mortality and morbidity risk for girls
\citep{Whitworth2002,Bhalotra2008,Maitra2008,Jayachandran2011,Jayachandran2017a}.
Longer spacing between births then potentially lowers mortality, although there may
be counteracting effects if the longer spacing arises because of multiple abortions.
There has, so far, been no attempt to examine which mechanism dominates.

% Finally, birth spacing has long served as a measure of son preference, with strong
% son preference associated with shorter birth intervals, the fewer sons a family has 
% \citep{ben-porath76b,Leung1988}.
% Now, longer birth intervals in the absence of sons may instead arise because couples 
% with a strong preference for sons use sex selection, possibly to the point where we 
% observe a reversal of the traditional spacing pattern.
% Despite a large literature on son preference and birth spacing, no prior research has,
% however, examined how sex selection affects spacing.

I examine how birth spacing in India has changed over time and across 
groups with the introduction of sex selection.
Using data from the four National Family and Health Surveys (NFHS), I apply
a competing risk hazard model to the birth histories of Hindu women, covering
the period 1972 to 2016.
The competing risk hazard model directly incorporates the effects of 
sex-selective abortions on birth intervals
\emph{and} 
the likelihood of a son, 
but also works in cases without prenatal sex selection.

@@

Third, to what extent does TFR reflect cohort fertility?
The empirical model allows me to predict fertility behavior taking into
account both the likelihood of parity progression and having a child of a specific sex.
I compare those predictions with fertility rate estimates based on the same data to
show how the increases in birth spacing have led to a substantial overestimate of how 
fast fertility has declined. [TK too imprecise? underestimate of the level, rather?]

@@

To set the stage for the subsequent analyses, I first describe how female education
and labor force participation has changed.
Both are potentially important factors in the birth spacing decision, as I discuss
in the next section.

@@

Female education is higher in urban areas than in rural areas.
Around sixty percent of urban women born in the 1930s had no education, 25 percent had
between one and seven years, about ten percent had eight or more years of education, and 
only five percent has twelve years or more.
The proportions with no education and one to seven years have both declined to just below 
ten percent for the latest cohort.
Although the proportion of women with eight to eleven years of education has increased
to about twenty, most of the increase in urban female education has come from the 
twelve plus group, which now account for more than half of all urban women.

@@

The NFHS data show a U-shaped relationship between education and working for married
women, with the highest percent working for women with either no education 
or with twelve or more years and the lowest for women with eight to eleven years of 
education.%
\footnote{
See Appendix Figures \ref{fig:work_by_survey} through \ref{fig:work_family_by_survey}. 
}
Women are more likely to report working if they live in rural than urban areas 
and the older they are.
For married women there is relatively little change in the percentage currently working 
over time,
but all education and age groups have become substantially more likely to work for
a family member.

The increases in female educational attainment imply that access to education has 
expanded beyond the higher castes. 
One possible effect of the rapid expansion in female education, and the 
associated change in the composition of better-educated women, is that the behavior
of the better-educated would change. 
However, ``Sanskritization'' implies that as lower castes females gain access to 
education and their husbands' income increases, the women adopt higher caste 
norms such as stronger son preference and a retraction from the formal labor 
market \citep{Srinivas1956,Chen1995,Abraham2013,Chatterjee2018}.
The low and declining female labor force participation is a first indication that this
process still operates.

@@

The standard economic argument for why increasing female education has predominantly
been associated with shorter spacing is that parents incur time costs when they have 
children \citep{Hotz1997,schultz97}.
Specifically, if having children require the mother to reduce her market work, parents 
can lower the cost of children by shortening birth spacing to 
take advantage of economies of scale in childrearing \citep{Vijverberg1982}.

The low and declining female labor force participation, especially for younger women, 
suggests, however, that families face little incentive to space children more closely 
together for economic reasons.
This may happen if household income has increased so substantially that the income effect 
dominates any substitution effect.
Two findings speak for this effect.
First, although real wages for both men and women have almost doubled between 1987 and 
2011, the mean male wage is still close to 70 percent higher than the female wage 
\citep{Klasen2015,Bhargava2018}.
Second, women’s labor supply appear to be more negatively elastic to their husbands' 
wages than it is positively elastic to own wages \citep{Bhargava2018}.

The combination of rising incomes and continued son preference may even lead to longer 
spacing than what we would expect from the standard income effect alone.
As women's education increases, their productivity at home also increases, especially in 
the production of offspring human capital.
With relatively more boys born because of increased access to sex-selective 
abortions and an increasing potential income for (male) offspring, demand for 
better-educated women can increase, even if they do not participate in the labor market 
\citep{Behrman1999}.
The result would be longer birth intervals if parental attention per child results in 
higher child ``quality'' \citep{Zajonc1975,Zajonc1976,Razin1980}.
The evidence on spacing's effect on child quality measures such as IQ 
and education is, however, mixed for developed countries and non-existing 
for developing countries
\citep{Powell1993,Pettersson-Lidbom2009,Buckles2012,Barclay2017}.
Nevertheless, for one crucial aspect of child quality---health and mortality---longer 
spacing does lead to better outcomes, although this relationship weakens with maternal 
education \citep{Whitworth2002,Conde-Agudelo2012,Molitoris2019}.

% While one might expect that increasing income and female education would also lengthen
% average birth spacing through better access and knowledge of contraception, the empirical 
% relationship between contraception use and birth spacing is ambiguous 
% \citep{Tulasidhar1993,Whitworth2002,Bhalotra2008,Yeakey2009,Kim2010,Soest2018}.
% Furthermore, the ability to successfully use ``low efficacy'' contraceptive methods, 
% such as the rhythm method, is increasing in education, while there is no difference 
% across education levels when it comes to modern contraceptives such as pills or 
% injectables \citep{Rosenzweig1989}.

With the introduction of sex selection, parents can avoid the birth of girls, 
although at the cost of the procedure and having to wait longer until the next child 
for each aborted pregnancy.
Theory suggests that sex selection increases with lower desired fertility, and, for a 
given desired number of children, the higher the parity \citep{Portner2015b}.
The increased use of sex selection in India with education and in urban
areas is consistent with a lower desired fertility for both groups
\citep{das_gupta97,retherford03b,Guilmoto2009a,Portner2015b,Jayachandran2017}.
Better-educated women also tend to live in households with higher income, which better 
enables them to access prenatal sex determination and lowers the relative costs of using 
sex selection and have long birth intervals.
We may even observe a ``delay'' effect, where parents conceive later because they know 
that short spacing is detrimental to the next child's health, and that the next child is 
more likely to be a boy because of access to sex selection.

In summary, with substantial increases in husbands' income and a declining female labor 
force participation, I expect a push toward longer birth spacing over time, independent
of education levels, based on the income effects and the effects of spacing
on child outcomes.
As desired fertility decreases with increasing education, I, furthermore, expect 
birth spacing to increase the most among the better-educated because of their 
use of sex selection.%
\footnote{
Working in the opposite direction is that women with more education can space
their children closer together without substantially increasing child mortality risk,
although compared with those factors this effect is likely to be relatively small.
}
Even with the large increase in the number of better-educated women, "Sanskritization" 
imply that the changing composition does not substantially change this group's use of 
sex selection.

@@

The standard approach in the birth spacing literature is to use proportional hazard
models with a single exit, the birth of a child.%
\footnote{
See \citet{Sheps1970} and \citet{Newman1984} for early discussions of why 
hazard models are the preferred way to deal with the censoring of birth
intervals.
}
There are two problems with the standard approach in this setting.
First, and most importantly, the introduction of sex selection means that the sex of the 
next child is no longer necessarily random and that parents' choices will impact the 
spacing to the birth of a girl or a boy differently.
I, therefore, use a competing risk set-up, which can capture both the non-randomness of
the birth outcome and the differential spacing.%
\footnote{
\cite{Merli2000} used a discrete hazard model to examine whether 
there were under-reporting of births in rural China, although they 
estimated separate waiting time regressions for boys and girls.
}

@@

% The distribution of spacing is captured by the survival curve, which shows the probability 
% of not having had a birth yet by spell duration, for a given set of explanatory variables.
% The survival curve at time $t$ is 
% \begin{equation}
% \label{eq:survival}
% S_{t} 
% = 
% \prod_{d=1}^t
% \left(
% \frac{ 1 }
% {1 + \sum_{l=2}^2 \exp(D_j(t) + \alpha_{ld}'\mathbf{Z}_{kd} + \beta_l'\mathbf{X}_{k})}
% \right).
% \end{equation}
% 
% Because the probability of ever having a next birth varies across groups, a direct 
% comparison of standard survival curves tells us little about how the spread of sex 
% selection affects birth spacing across groups.
% I, therefore, condition on the predicted likelihood of parity progression when examining 
% birth spacing measures, such as the average duration to a birth.
% The reliability of this approach depends on whether the spell length covered is 
% sufficiently long that few women are likely to give birth after the spell cut-off.
% I discuss the choice of spell length below.%
% \footnote{
% It is important to note that the approach is not the same as merely calculating 
% the birth spacing measures for women who already have a given 
% parity child in the survey because that number does not take into account
% the censoring of spells that will eventually lead to a birth.
% }
% In addition, I present graphs of survival curves conditional on parity progression,
% which therefore begin at 100\% and end at 0\%.

Interpretation of the model coefficients is challenging \citep{thomas96}.
It is, however, possible to calculate the predicted probabilities of 
having a boy, $b$, and of having a girl, $g$, in period $t$, conditional on 
a set of explanatory variables and not having had a child before that period.
% , as
% \begin{align}
% P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt}, t ) 
% & =  
% \frac{ \exp(D_j(t) + \alpha_{1t}' \mathbf{Z}_{kt} + \beta_1' \mathbf{X}_{k} )}
% {1 + \sum_{l=1}^2 \exp(D_j(t) + \alpha_{lt} ' \mathbf{Z}_{kt} + \beta_l ' \mathbf{X}_{k})}
% \label{eq:probability_boy} \\
% P(g_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t ) 
% & =  
% \frac{ \exp(D_j(t) + \alpha_{2t}'\mathbf{Z}_{kt} + \beta_2'\mathbf{X}_{k} )}
% {1 + \sum_{l=2}^2 \exp(D_j(t) + \alpha_{lt}'\mathbf{Z}_{kt} + \beta_l'\mathbf{X}_{k})}
% \label{eq:probability_girl}
% \end{align}
It is then straightforward to calculate the estimated percentage of children born that 
are boys, $\hat{Y}$, at each $t$:  
\begin{equation}
\hat{Y}_t 
= 
\frac{ P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )}
{ P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t) + P(g_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )} 
\times 100.
\label{eq:probability_son}
\end{equation}
Combining the percentage boys and the likelihood of exiting the spell 
across all $t$ gives the predicted percent boys born over the entire spell.

@@

In addition to the large number of women surveyed and the long period covered, a
benefit of the NFHS is that surveys enumerators pay careful attention to spacing between 
births and probe for ``missed'' births.
Nevertheless, systematic recall error, where the likelihood of reporting a deceased 
child depends on the sex of the child, is still a potentially substantial problem and
will bias both spacing and sex ratios.
As I show in the online Appendix, recall error is heavily dependent on how long ago a 
woman was married, and I, therefore, drop women married 22 years or more.
The final sample consists of 
\input{../tables/num_women.tex}
% [number] women, with [number] parity one through four births.

Direct information on the use of sex selection is not available, so I compare different 
periods as a proxy, based on the changes in access and legality of prenatal sex 
determination in India.
Abortion has been legal in India since 1971.
Reports of sex determination appeared around 1982--83, and the number of clinics 
quickly increased \citep{Sudha1999,bhat06,Grover2006}.
In 1994, the Prenatal Diagnostic Techniques (PNDT) Act made determining and communicating 
the sex of a fetus illegal.%
\footnote{
Details about the act are at \href{http://pndt.gov.in/}{http://pndt.gov.in/}.
There is little evidence that the ban significantly affected sex ratios \citep{Das-Gupta2016}.
}
Although the use of sex selection increased even after 1994, some researchers argue that 
we have passed a turning point in its use \citep{Das_Gupta2009,Diamond-Smith2015}.
I, therefore, use four periods: 1972--1984, 1985--1994, 1995--2004, and 2005--2016.
The allocation of spells into periods is determined by when conception, and therefore 
decisions on sex selection, can begin. 
This allocation mechanism means that some spells cover two periods, which may bias 
downward the differences between the periods.

I focus here on the second through fourth spells, that is, on the intervals from the first 
birth until at most the fourth birth.
I exclude the spell from marriage to the first birth because of data issues.
In NFHS-1 and NFHS-2 the only information on marriage is the age at marriage, and 
all first spell durations are therefore imputed.
In NFHS-3 about a third of respondent did not provide complete information on the
timing of marriage and had the length of their first spell imputed using the age at 
marriage.
The imputation results in large numbers of women with spells that are less than nine 
months and makes a comparison of spell duration over time subject to an 
excessive amount of noise.

The spells all begin nine months after the previous birth, which is the earliest we should 
expect to observe a new birth.
A spell continues until either a child is born or censoring occurs.
Censoring can happen for three reasons:
the survey takes place;
sterilization of the woman or her husband;
or too few births are observed for the method to work.
For all spells censoring is set at 96 months (eight years) after a birth can occur.
With these cut-offs, less than one percent of observed births occur after
the spell cut-off.

% %
% \footnote{
% \input{../tables/num_missed.tex}
% }

@@

Women with different education levels have different hazard profiles
\citep{Whitworth2002,Bhalotra2008,Kim2010}.
Furthermore, as discussed above, the use of sex selection increases with education, 
either because of the associated lower fertility or because higher income enables them 
to access and afford prenatal sex determination
I divide women into four groups based on education attainment: no
education, one to seven years of education, eight to eleven years of education,
and twelve and more years of education.
The models are estimated separately for each education level to reduce the potential 
problem of including other variables as proportional.

@@

\section{How Birth Spacing has Changed\label{sec:results}}

% [How to read tables]
I estimate the model for each spell/education/period subsample, the results of which I use 
to predict average birth spacing, sex ratio, and the probability of having a birth in that 
spell.
Figures \ref{fig:bootstrap_low} through \ref{fig:bootstrap_highest}
show these predicted outcomes for the four education levels separated by the area of 
residence.%
\footnote{
For legibility, the graphs do not show standard errors.
The graphed values, with standard errors, are shown in Appendix Tables 
\ref{tab:avg_sex_ratio_low} through \ref{tab:avg_sex_ratio_highest}.
}

The predicted parity progression probability is the likelihood of a birth by 96 months 
after the beginning of the spell.
To find the expected average duration I first calculate, for each woman, the probability 
of giving birth in each period $t$, which I use as weights to calculated her expected 
duration. 
I then average the individual expected durations across women using the parity progression
probabilities as weights.

The predicted sex ratio captures the percent boys to women in the sample when 
childbearing for that spell is over.
A woman's predicted sex ratio is the weighted average of the predicted percentage boys 
over each $t$ in the spell, calculated using equation (\ref{eq:probability_son}), using
the probability of giving birth at time $t$ as weights.%
\footnote{
If T=2 with the estimated percent boys 54\% and 66\% and the likelihoods of giving 
birth 20\% and 40\%, the percent boys is $\frac{54*0.2+66*0.4}{0.2+0.4} = 62$.
}
The predicted sex ratio shown is the weighted average of the individual predicted sex 
ratios, using the parity progression probabilities as weights.
For comparison, the graphs also show the natural sex ratio, approximately 51.2\% 
\citep{ben-porath76b,jacobsen99,Portner2015b}.

@@

% \captionsetup[subfigure]{position=top,captionskip=-1pt,farskip=-0.5pt}

\captionsetup[subfigure]{captionskip=-1pt,farskip=-0.5pt}
